---
title: "Machine Learning Homework 4.2"
author: "Josh Elder and Curt Ginder"
header-includes: null
output:
  word_document: default
  pdf_document: default
geometry: left=2.5cm,right=2.5cm,top=.5cm,bottom=1.5cm
---
First, we read in libraries including the "recommenderlab" needed for this assignment.
```{r include=FALSE}
PackageList=c("MASS", "ISLR","animation","ElemStatLearn","glmnet","textir","nnet","methods","statmod","stats",
              "graphics","RCurl","jsonlite","tools","utils","data.table","gbm","ggplot2","randomForest",
              "tree","class","kknn","e1071","data.table","R.utils","recommenderlab","h2o")
NewPackages=PackageList[!(PackageList %in% 
                            installed.packages()[,"Package"])]
if(length(NewPackages)) install.packages(NewPackages)
lapply(PackageList,require,character.only=TRUE)#array function
```

Then, we read in the JSON file that contains video game ratings: 
```{r include=FALSE}
fileConnection <- gzcon(file("videoGames.json.gz", "rb"))
InputData = stream_in(fileConnection)
```

and convert it into a "realRatingMatrix" to use with the recommenderlab package

```{r include=FALSE}
ratingData = as(InputData[c("reviewerID", "itemID", "rating")], "realRatingMatrix")

```
We then remove users with fewer than 2 video games rated or video games with fewer than 3 ratings total.

```{r}

# we keep users that have rated more than 2 video games
ratingData = ratingData[rowCounts(ratingData) > 2,]

# we will focus only on popular video games that have 
# been rated by more than 3 times
ratingData = ratingData[,colCounts(ratingData) > 3]

# we are left with this many users and items
dim(ratingData)
```

We are left with 2656 users and 632 video games with ratings.

We then explore features of the recommendarlab:

```{r}
# example on how to recommend using Popular method
r = Recommender(ratingData, method="POPULAR")

# recommend 5 items to user it row 13
rec = predict(r, ratingData[13, ], type="topNList", n=5)
as(rec, "list")

# predict ratings 
rec = predict(r, ratingData[13, ], type="ratings")
#as(rec, "matrix")
```

## Task 1 & 2 - Find the User And Video Game with the most amount of reviews
We then explore the data to identify users with the most ratings and the video game with the most ratings.
```{r}
# Find user with most ratings
most.ratings.user <- names(which.max(rowCounts(ratingData)))
most.ratings.user.reviews <- max(rowCounts(ratingData))
cat(c(most.ratings.user,"has rated the most video games with",most.ratings.user.reviews,"reviews","\n"))

most.rated.vg <- names(which.max(colCounts(ratingData)))
most.rated.vg.reviews <- max(colCounts(ratingData))
cat(c(most.rated.vg,"is the most rated video game with",most.rated.vg.reviews,"reviews"))
```

It looks like User "U584295664" has given 53 ratings, while Video Game "I760611623" has received 200 reviews.


## Task 3 Find User that is the most similar to "U141954350"
When we initially set out on this problem we attempted to identify similar users to U141954350 using the realRatingData matrix created above. However, upon further exploration of the data, we realized that the user of interest is a tough critic. Among all ratings by this user, they have ranked every game a "1"

```{r}
print(InputData[InputData$reviewerID == "U141954350",c(1:2)])
```

This causes a problem when we look at the types of similarities and recommendations we can offer this user. We decided that the presence of a rating, regardless of score, might offer some insight into providing a recommendation. Because this user is reviewing the game, they likely have purchased and used the game. Their selection of individual games to play and rate (albeit, a score of a "1"), might reveal their true preference of other games they would like to purchase.

Because of this finding, we decided to convert our ratingData matrix to a binary matrix, where a 1 equals a purchase and rating, regardless of score, of a given item.

```{r}
# Find Similar User to U141954350
binary.InputData <- InputData[c("reviewerID","itemID","rating")]
binary.InputData$rating <- 1

binary.ratingData <- as(binary.InputData[c("reviewerID", "itemID", "rating")], "binaryRatingMatrix")
# keep user that have rated more than 2 video games
binary.ratingData = binary.ratingData[rowCounts(binary.ratingData) > 2,]

# we will focus only on popular video games that have 
# been rated by more than 3 times
binary.ratingData = binary.ratingData[,colCounts(binary.ratingData) > 3]
```

We then separated the rating data into two separate binaryRatingMatrices, one for the user, and one for all other users.

```{r}

currentUser <- "U141954350"

bin.otherUsers <- binary.ratingData[!(rownames(binary.ratingData) %in% c(currentUser)),]
bin.currentUserRatings = binary.ratingData[(rownames(binary.ratingData) %in% c(currentUser)),]

compare_1 = similarity(bin.currentUserRatings,bin.otherUsers,method='jaccard',which = 'users')
compare_1_mat=as(compare_1,'matrix')
```

After doing this, we selected the user with the highest jaccard-calculated similarity.

```{r}
matchedUser = colnames(compare_1_mat)[compare_1_mat == max(compare_1_mat)]
cat('The user that is most similar to',currentUser, 'is',matchedUser,'\n')
```

Looking into this matched user, we can evaluate which games they also reviewed. Below displays the actual ratings (as compared to the binary ratings)

```{r}
InputData[InputData$reviewerID == matchedUser,c(1:2)]

```

Interestingly, it looks like this user also gave ratings of "1" to every game they reviewed. Additionally, we noticed that the games the matched user reviewed were also reviewed by our initial user of interest, giving us a Jaccard Similarity of "1".

## Task 4 - Recommend Movies to User U141954350
We decided to use a User-Based Collaborative Filter to recommend a game that this user has not played. We decided to use UBCF using the binary ratingMatrix for similar reasons as discussed above. Our hypothesis is that users who play and rate similar types of games can give recommendations for other games that the user of interest has not personally played. We trained the UBCF recommendation system using all other users without the user of interest.

If our current user had provided a more diverse set of rating ranges, we could have pursued other recommendation strategies, including an item based collaborative filter.

```{r}
recommenderRegistry$get_entry('UBCF', dataType='binaryRatingMatrix')

ubcf_rec = Recommender(
  data=bin.otherUsers,
  method='UBCF',           # User-Based Collaborative Filtering
  parameter=list(
    normalize='center',    # normalizing by subtracting average rating per user;
                           # note that we don't scale by standard deviations here;
                           # we are assuming people rate on the same scale but have
                           # different biases
    method='Jaccard',      # use Pearson correlation
    nn=30                  # number of Nearest Neighbors for calibration
  ))

ubcf_rec
```
```{r}

ubcf_pred = predict(ubcf_rec,bin.currentUserRatings,type='topNList',n=5)

predictedTop=as(ubcf_pred,'list')[1]
predictedTop.List <- as.vector(unlist(predictedTop))

cat(c("We predict that User U141954350 would also play and review:\n",predictedTop.List))
```
